Research Notes

### 2018-07-17 writups/notes.tex
update writeups, clean up dictionary merging

@@

Due to the structure of \lstinline{Dedalus}, we must use Chebyshev polynomials
in the vertical direction (to be able to enforce BCs). But as discussed in
\autoref{s:chebyshev}, the grid spacing of the Chebyshev polynomials near
boundaries goes as $1/N^2$, which means that interface forcing necessitates an
extremely strict Courant condition to capture evolution near the interface; this
is the explanation for the divergences we were seeing in earlier sections.

@@

\subsection{1D Wave Equation}

@@

>>>>

\chapter{Simulation Setup}

We describe now some concerns that arise in setting up the simulation.

@@

Due to the structure of \lstinline{Dedalus}, we must use Chebyshev polynomials
in the vertical direction (to be able to enforce BCs). But as discussed in
\autoref{s:chebyshev}, the grid spacing of the Chebyshev polynomials near
boundaries goes as $1/N^2$, which means that interface forcing necessitates an
extremely strict Courant condition to capture evolution near the interface; this
is the explanation for the divergences we were seeing in earlier sections. In
fact, in the literature, the spectral CFL condition meets a $\Delta t_{\max}
\sim 1/N^2$ scaling thanks to poor bounds near the boundary.

@@

\subsection{1D Wave Equation (Deprecated)}

@@

\subsection{Narrow-band forcing}

Instead of doing this on-frequency forcing, we choose instead to use a narrow
forcing zone $f(x, t) \sim N_{\sigma}(z)e^{it}$ where $\sigma$ is not large.
Specifically, we need $\sigma$ to be well-resolved spectrally, but we need
$\tilde{f}(k_z, t)$ not to be small. Thus, since $1/k_z$ must be resolved by the
simulation, choosing $\sigma = 1/k_z$ is a natural choice.

Note that this implies that waves will be excited going both upwards and
downwards, as $\tilde{f}(k_z) = \tilde{f}(-k_z)$. We can however place the
forcing zone near the bottom boundary so that downwards-propagating waves will
be attenuated by damping zones.

\section{Timestepping}

\subsection{Signatures of Insufficient Temporal Resolution}

When running these simulations, it can occasionally be seen that the waves do
not grow in amplitude like $e^{z/2H}$, such that the energy and flux are not
constant (or the waves will grow too quickly; this is dependent on the exact
timestepping used). This is caused by using an overly-long timestep, as can be
seen with the below illustration.

We study the 1D advection equation $\psi_t = c\psi_x$ corresponding to
rightwards propagating waves. Consider first order implicit Euler timestepping
for wave mode $\psi_x = ik\psi$, then
\begin{equation}
    \psi_{i + 1} - \psi_i = ick\Delta t\psi_{i + 1}.
\end{equation}
Solving yields
\begin{align}
    \psi_{i + 1} &= \p*{1 - ick\Delta t}^{-1}\psi_i,\nonumber\\
        &= \p*{1 - i\omega \Delta t}^{-1}\psi_i.
\end{align}
If we then want to step forwards for some total time $T$, with $N \Delta t = T$
the number of timesteps taken, then we approximate
\begin{align}
    \psi(t_0 + T) \approx \p*{1 - i\omega \frac{T}{N}}^{-N}\psi(t_0).
\end{align}
It is worth noting that in the $N \to \infty$ limit we recover $\psi(t_0 + T) =
e^{i\omega T}\psi(t_0)$.

However, consider if $\omega \Delta t \lesssim 1$, then we should see the wave
amplitude decrease over time like
\begin{equation}
    \abs*{\psi(t_0 + T)} \approx \p*{1 + \omega^2 \Delta t^2}^{-N/2}\psi(t_0).
\end{equation}
Thus, since we have an implicit first order scheme, the wave amplitude
\emph{decreases} over time spuriously. An explicit method would produce an
increase, but \lstinline{Dedalus} uses implicit timestepping. The exact rate of
decrease depends on the timestepper used, but it is clear that $\omega \Delta t
\ll 1$ is necessary to avoid this spurious decrease (in numerics language, ``the
oscillation must be well-resolved'').
